---
title: "p8105_hw3_ss7332"
author: "Sean Sorek"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      message = F,
                      warning = F,
                      fig.align = "center")
library(tidyverse)

theme_set(theme_bw() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

# Problem 2

```{r}
covar <- read_csv("data/nhanes_covar.csv", skip = 4) |> na.omit() |>
  mutate(
    sex = factor(sex, 1:2, c("male", "female")),
    education = factor(education, 1:3, 
                       c("Less Than High School",
                         "High School",
                         "More Than High School"))) |>
  filter(age >= 21)
accel <- read_csv("data/nhanes_accel.csv") |>
  pivot_longer(
    starts_with("min"),
    names_prefix = "min",
    names_to = "minute",
    values_to = "MIMS") |>
  mutate(minute = as.integer(minute))

full_ac <- inner_join(covar, accel, by = join_by(SEQN)) |> glimpse()

full_ac |> select(SEQN, sex, education) |> distinct() |>
  group_by(sex, education) |> summarize(count = n()) |>
  pivot_wider(names_from = sex, values_from = count) |> knitr::kable()
```

```{r}
full_ac |> select(SEQN, sex, education) |> distinct() |> 
  ggplot(aes(x = education, fill = sex)) + geom_histogram(stat = "count", position = "dodge") +
  labs(title = "Education Dist. by Gender") + theme_bw()
```

There are more men with only High School education than women in the dataset. The other two groups have near equal representation.

```{r}
full_ac |>
  group_by(SEQN) |>
  summarise(total_activity = sum(MIMS)) |>
  inner_join(y = full_ac) |>
  select(SEQN, total_activity, age, sex, education) |>
  distinct() |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point() + geom_smooth() + facet_wrap(vars(education)) +
  labs(title = "Activity Level by Sex and Education",
       x = "Age", 
       y = "Total Activity Over 24 Hours")
```

Total activity decreases as subjects increase in age. The negative effect of age appears to be stronger in men than in women for those with less than a high school education. For those with a high school education or more, women tend to have slightly higher total activity levels.

```{r, fig.height=7, fig.width= 7}
full_ac |>
  ggplot(aes(x = minute, y = MIMS,  color = sex)) +
  geom_line(aes(group = SEQN),alpha = .2) + geom_smooth() + facet_wrap(vars(education)) +
  labs(title = "Activity Level per Minute by Sex and Education", x = "Minute", y = "Activity (MIMS)")
```

This is a much messier visualization than the previous. The only trend discernable to me is a slightly higher activity rate for women that have more than a high school education.

# Problem 3

```{r}
july_20 <- read_csv("data/July 2020 Citi.csv", col_types = "cffdfff") |>
  mutate(month = "July", year = 2020)
july_24 <- read_csv("data/July 2024 Citi.csv", col_types = "cffdfff")|>
  mutate(month = "July", year = 2024)
jan_20 <- read_csv("data/Jan 2020 Citi.csv", col_types = "cffdfff") |>
  mutate(month = "January", year = 2020)
jan_24 <- read_csv("data/Jan 2024 Citi.csv", col_types = "cffdfff") |>
  mutate(month = "January", year = 2024)
citibike <- bind_rows(july_20, 
                      july_24,
                      jan_20,
                      jan_24) |>
  mutate(month = as.factor(month),
         weekdays = factor(weekdays,
                           levels= c("Sunday",
                                  "Monday",
                                  "Tuesday",
                                  "Wednesday",
                                  "Thursday",
                                  "Friday",
                                  "Saturday"))) |> glimpse()
```

This dataset contains individual data for `r nrow(citibike)` rides of the Citi Bike bike sharing system in NYC. It represents data from both January and July of 2020 and 2024. For each ride, the dataset shows the duration, start/end position, weekday, bike type, and membership status of the rider.

```{r}
citibike |> select(month, year, member_casual) |>
  group_by(month, year, member_casual) |>
  summarize("total_rides" = n()) |>
  pivot_wider(names_from = member_casual, values_from = total_rides) |>
  knitr::kable()
```

The number of citi bike users members has nearly doubled from 2020 to 2024. This effect is especially noticable in July.

```{r}
citibike |> filter(month == "July", year == 2024) |>
  select(start_station_name) |>
  group_by(start_station_name) |>
  summarize("Rides" = n()) |>
  arrange(desc(Rides)) |> slice(1:5) |> knitr::kable()
```

The most popular stations are in the western half of downtown Manhattan, mostly in Chelsea.

```{r}
citibike |>
  group_by(weekdays, month, year) |>
  summarise(median_duration = median(duration)) |>
  ggplot(aes(x = weekdays, y = median_duration, color = month)) +
  geom_line(aes(group = month), linewidth = 1) + facet_wrap(vars(year)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(title = "Median Duration Over a Week")
```

Median duration in July decreased across the entire week from 2020 to 2024. The effect of the weekend on ride duration appears to be mitigated in both months.

```{r}
citibike |> filter(year == 2024) |> ggplot(aes(x = duration, fill = member_casual)) + facet_wrap(vars(month, rideable_type)) + geom_histogram() + scale_x_log10()
```

The number of casual riders appeared to rise significantly compared to member riders. There is also a spike in electric bike usage.